---
title: "Kuhlanapo Gage Data Figures for Reports"
author: "Skyler Lewis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup}
knitr::opts_chunk$set(
	fig.height = 4,
	fig.width = 6.5,
	message = FALSE,
	warning = TRUE,
	dpi = 300
)
library(tidyverse)
library(janitor)
library(patchwork)

theme_custom <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "black", linewidth = 0.3),
      axis.ticks = element_line(color = "black", linewidth = 0.3),
      axis.title = element_text(size = rel(0.9)),
      plot.title.position = "plot"
    )
}
theme_set(theme_custom())

source(here::here("global.R"))

# As published on USGS lake level gage 11450000
navd88_to_rumsey_usgs <- function(x) x - 1320.74
rumsey_to_navd88_usgs <- function(x) x + 1320.74
```

## Import Data

Import time series data

```{r data_import}
ts_data <-  readRDS(here::here("data", "gage_data.Rds")) 
ll_data <- readRDS(here::here("data/usgs_lake_level_11450000.rds"))
precip_data <- readRDS(here::here("data/precip_ts.rds"))
```

Apply data cleaning via `build_df_pivot()` (defined in `global.R`), shared
with `app.R` so the two can never drift out of sync.

```{r data_cleaning}
df_pivot <- build_df_pivot(ts_data, ll_data)
```

## Groundwater Study

```{r gw_elev, fig.width=6.5, fig.height=4, dpi=300}
plt_gw_elev <-
  df_pivot |>
  filter(category == "Piezometer") |>
  filter(timestamp >= ymd("2025-12-05")) |>
  filter(timestamp < Sys.Date()) |>
  ggplot(aes(x = timestamp, y = gwe_ft_navd88)) +
  geom_line(data = ll_data |> filter(timestamp < Sys.Date()),
            aes(y = value, linetype = "Lake Level")) +
  geom_line(aes(color = code)) +
  geom_hline(aes(yintercept = rumsey_to_navd88_usgs(7.56),
                 linetype = "Full Lake")) +
  scale_x_datetime(name = "",
                   date_breaks = "1 month",
                   expand = c(0, 0)) +
  scale_y_continuous(name = "Elevation (ft NAVD88)",
                     breaks = scales::breaks_width(1),
                     sec.axis = sec_axis(name = "Elevation (ft Rumsey)", 
                                         transform = ~ navd88_to_rumsey_usgs(.),
                                         breaks = scales::breaks_width(1))) +
  scale_color_manual(name = "Piezometers",
                     values = piezo_colors) +
  scale_linetype_manual(name = "",
                        values = c("Lake Level" = "solid",
                                   "Full Lake" = "dotted")) +
  theme(panel.grid.major = element_line(),
        axis.text.x.bottom = element_text(angle = 45, hjust=1))

print(plt_gw_elev)
```

```{r gw_depth, fig.width=6.5, fig.height=4, dpi=300}
plt_gw_depth <-
  df_pivot |>
  filter(category == "Piezometer") |>
  filter(timestamp >= ymd("2025-12-05")) |>
  filter(timestamp < Sys.Date()) |>
  ggplot(aes(x = timestamp, y = gw_depth_ft)) +
  geom_line(aes(color = code)) +
  geom_hline(aes(yintercept = 0,
                 linetype = "Ground")) +
  scale_x_datetime(name = "",
                   date_breaks = "1 month",
                   expand = c(0, 0)) +
  scale_y_reverse(name = "Depth Below Ground Surface (ft)",
                  breaks = scales::breaks_width(1)) +
  scale_color_manual(name = "Piezometers",
                     values = piezo_colors) +
  scale_linetype_manual(name = "",
                        values = c("Ground" = "dashed")) +
  theme(panel.grid.major = element_line(),
        axis.text.x.bottom = element_text(angle = 45, hjust=1))

print(plt_gw_depth)
```

```{r gw_precip, fig.width=6.5, fig.height=4, dpi=300}
plt_gw_precip <-
  precip_data |>
  filter(site == "KPD") |>
  filter(timestamp >= ymd("2025-12-05")) |>
  filter(timestamp < Sys.Date()) |>
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_rect(aes(xmin = timestamp,
                xmax = timestamp + hours(1),
                ymin = 0,
                ymax = precip_in)) +
  scale_x_datetime(name = "",
                 date_breaks = "1 month",
                 expand = c(0, 0)) +
  scale_y_continuous(name = "Precipitation (in)",
                     breaks = scales::breaks_width(0.1),
                     expand = c(0, NA)) +
  theme(panel.grid.major = element_line(),
        axis.text.x.bottom = element_text(angle = 45, hjust=1))

print(plt_gw_precip)
```

```{r gw_combined, fig.width=6.5, fig.height=8.2, dpi=300}
(plt_gw_precip / plt_gw_depth / plt_gw_elev) +
  plot_layout(heights = c(1, 2, 2), guides = "collect", axes = "collect_x")
```

## Surface Water

```{r sw_depth, fig.width=6.5, fig.height=4, dpi=300}
plt_sw_depth <-
  df_pivot |>
  filter(category == "Stage Gage") |>
  filter(timestamp >= ymd("2025-12-05")) |>
  filter(timestamp < Sys.Date()) |>
  ggplot(aes(x = timestamp, y = depth)) +
  geom_line(aes(color = code)) +
  scale_x_datetime(name = "",
                   date_breaks = "1 month",
                   expand = c(0, 0)) +
  scale_y_continuous(name = "Water Depth (ft)",
                     breaks = scales::breaks_width(1),
                     expand = c(0, NA)) +
  scale_color_manual(name = "Stage Gages",
                     values = site_colors) +
  theme(panel.grid.major = element_line(),
        axis.text.x.bottom = element_text(angle = 45, hjust=1))

print(plt_sw_depth)
```

```{r sw_elev, fig.width=6.5, fig.height=4, dpi=300}
plt_sw_elev <-
  df_pivot |>
  filter(category == "Stage Gage") |>
  filter(timestamp >= ymd("2025-12-05")) |>
  filter(timestamp < Sys.Date()) |>
  ggplot(aes(x = timestamp, y = wse_ft_navd88)) +
  geom_line(data = ll_data |> filter(timestamp < Sys.Date()),
            aes(y = value, linetype = "Lake Level")) +
  geom_line(aes(color = code)) +
  geom_hline(aes(yintercept = rumsey_to_navd88_usgs(7.56),
                 linetype = "Full Lake")) +
  scale_x_datetime(name = "",
                   date_breaks = "1 month",
                   expand = c(0, 0)) +
  scale_y_continuous(name = "Elevation (ft NAVD88)",
                     breaks = scales::breaks_width(1),
                     sec.axis = sec_axis(name = "Elevation (ft Rumsey)", 
                                         transform = ~ navd88_to_rumsey_usgs(.),
                                         breaks = scales::breaks_width(1))) +
  scale_color_manual(name = "Stage Gages",
                     values = site_colors) +
  scale_linetype_manual(name = "",
                        values = c("Lake Level" = "solid",
                                   "Full Lake" = "dotted")) +
  theme(panel.grid.major = element_line(),
        axis.text.x.bottom = element_text(angle = 45, hjust=1))

print(plt_sw_elev)
```

```{r sw_precip, fig.width=6.5, fig.height=4, dpi=300}
plt_sw_precip <-
  precip_data |>
  filter(site == "UMC") |>
  filter(timestamp >= ymd("2025-12-05")) |>
  filter(timestamp < Sys.Date()) |>
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_rect(aes(xmin = timestamp,
                xmax = timestamp + hours(1),
                ymin = 0,
                ymax = precip_in)) +
  scale_x_datetime(name = "",
                   date_breaks = "1 month",
                   expand = c(0, 0)) +
  scale_y_continuous(name = "Precipitation (in)",
                     breaks = scales::breaks_width(0.1),
                     expand = c(0, NA)) +
  theme(panel.grid.major = element_line(),
        axis.text.x.bottom = element_text(angle = 45, hjust=1))

print(plt_sw_precip)
```

```{r sw_combined, fig.width=6.5, fig.height=8.2, dpi=300}
(plt_sw_precip / plt_sw_depth / plt_sw_elev) +
  plot_layout(heights = c(1, 2, 2), guides = "collect", axes = "collect_x")
```

## Groundwater Contour Export

Export the IDW-interpolated groundwater elevation surface for a single
timestamp as a GeoTIFF raster and a shapefile of contour lines. This reuses
`interpolate_idw_at_time()` and the precomputed `idw_precomputed.rds` grid
from `global.R` / `data-raw/prep_idw_grid.R` -- the same objects the "Piezo
Map" tab in `app.R` uses -- so the exported surface matches what the app
shows.

```{r gw_contour_export, message=FALSE}
library(terra)
library(sf)

idw_obj <- readRDS(here::here("data-raw", "idw_precomputed.rds"))
kuhlanapo_bnd <- readRDS(here::here("data", "kuhlanapo_bnd.rds"))

export_date <- as.Date("2026-01-01")
t0 <- as.POSIXct(paste(export_date, "00:00:00"), tz = "America/Los_Angeles")

gw_ts <- df_pivot |>
  filter(category == "Piezometer") |>
  transmute(code, timestamp, value = gwe_ft_navd88)

z_mat <- interpolate_idw_at_time(idw_obj, gw_ts, t0 = t0, return_matrix = TRUE)

stopifnot(
  "No piezometer readings close enough to this time to interpolate a map." =
    any(!is.na(z_mat))
)

# idw_obj$grid is expand.grid(x = gx, y = gy) (x fastest), and z_mat is filled
# from it byrow = FALSE, so z_mat[i, j] is the value at x = gx[i], y = gy[j].
# Reorient into raster row/col order: row 1 = max y (north), cols = increasing x.
gx <- idw_obj$x
gy <- idw_obj$y
dx <- diff(gx)[1]
dy <- diff(gy)[1]

r_mat <- t(z_mat)
r_mat <- r_mat[rev(seq_len(nrow(r_mat))), , drop = FALSE]

gw_raster <- rast(
  r_mat,
  extent = ext(min(gx) - dx / 2, max(gx) + dx / 2,
               min(gy) - dy / 2, max(gy) + dy / 2),
  # same CRS the piezometer coordinates were transformed to in prep_idw_grid.R
  crs = "EPSG:6418"
)

raster_path <- here::here("data-raw", paste0("gw_elevation_", export_date, ".tif"))
writeRaster(gw_raster, raster_path, overwrite = TRUE)

# contour lines at 1 ft intervals, matching the app's geom_contour(binwidth = 1)
z_range <- range(values(gw_raster, na.rm = TRUE))
contour_levels <- seq(floor(z_range[1]), ceiling(z_range[2]), by = 1)

gw_contours <- as.contour(gw_raster, levels = contour_levels) |>
  st_as_sf() |>
  rename(elev_ft_navd88 = level)

contour_path <- here::here("data-raw", paste0("gw_contour_lines_", export_date, ".shp"))
st_write(gw_contours, contour_path, delete_dsn = TRUE, quiet = TRUE)

cat("Raster written to:  ", raster_path, "\n")
cat("Contours written to:", contour_path, "\n")
```

```{r gw_contour_export_preview, fig.width=6.5, fig.height=6.5, dpi=300}
gw_raster_df <- as.data.frame(gw_raster, xy = TRUE) |> rename(value = 3)

ggplot() +
  geom_polygon(data = kuhlanapo_bnd, aes(x = X, y = Y), fill = "#eeeeee") +
  metR::geom_contour_fill(data = gw_raster_df, aes(x = x, y = y, z = value),
                          na.fill = TRUE, alpha = 0.6) +
  geom_sf(data = gw_contours, aes(color = elev_ft_navd88), linewidth = 0.8) +
  geom_point(data = idw_obj$sites, aes(x = x, y = y, fill = id), size = 10, shape = 21) +
  geom_text(data = idw_obj$sites, aes(x = x, y = y, label = substr(id, 4, 5)), size = 10 / .pt) +
  coord_sf(xlim = range(gx), ylim = range(gy)) +
  scale_fill_manual(name = "", values = piezo_colors, guide = "none") +
  scale_color_viridis_c(name = "ft NAVD88") +
  labs(title = paste("Groundwater Elevation --", export_date)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank())
```

